Which of my Transient Type Checks are not (Almost) Free?

### 2019-09-11 paper.tex
More changes

\setcopyright{acmlicensed}
\acmPrice{}
\acmDOI{10.1145/3358504.3361232}
\acmYear{2019}
\copyrightyear{2019}
\acmISBN{978-1-4503-6987-9/19/10}
\acmConference[VMIL '19]{Proceedings of the 11th ACM SIGPLAN International Workshop on Virtual Machines and Intermediate Languages}{October 22, 2019}{Athens, Greece}
\acmBooktitle{Proceedings of the 11th ACM SIGPLAN International Workshop on Virtual Machines and Intermediate Languages (VMIL '19), October 22, 2019, Athens, Greece}

@@

\section{Introduction}

Gradual typing aims to add static type annotations to dynamic languages, increasing
their safety while maintaining flexibility
\citep{GiladPluggable2004,Siek2006,XXXSiek2015}
--- or, complementarily,
to permit dynamic type annotations within static languages, increasing
flexibility whilst maintaining safety
\citep{AbadiTOPLAS1991}.

There is a spectrum of different approaches to gradual typing
\cite{kafka18,bensurvey18icfp}.
At one end is ``pluggable types'' (as in Strongtalk \cite{strongtalk}) or ``erasure
semantics'' (as in 
%% Two important approaches are optional\citep{}
%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
%% These are applied to dynamic languages to reap the benefits of typing, but
%% unfortunately also have limitations.
%With optional or pluggable approaches such as 
%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
%2018}{https://www.typescriptlang.org/}
TypeScript
\citep{typeScriptECOOP}), where
%
all types are erased before execution, limiting the benefit of
types to statically typed portions of programs, while not providing 
any for of run time checking.  In the middle there are
``transient'' or ``type-tag'' checks (as in Reticulated Python),
% and ``concrete'' checks as in Thorn
which offer first-order semantics, by checking
that an object's type constructor or supported methods match
% any explicit type annotations
any static types
that the object flows through.
%%%WRONG kjx
%, and checking only when data flows through those declarations
\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
Reticulated Python also supports alternative ``monotonic'' semantics
which mutate objects by narrowing their concrete types when passed
into contexts with more specific types.
On the other end of the spectrum are behavioural
type checks (as in Typed Racket \cite{typedScheme08,takikawa2012},
Gradualtalk \cite{gradualtalk14}),
and ``proxies'' (provided by Reticulated Python), which
support higher-order semantics, retaining
types until run 
time, and performing checks eagerly, thus detailed information
about type violations can be given as soon as possible via ``blame''
tracking \cite{blame2009,blameForAll2011}.
Finally, there is also ductile typing,
which dynamically interprets a static type system at run time
\cite{Ductile2011}.
%
%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
%  after concrete and before behavioural. Does it matter?}
%
% This said "any", but the ecoop paper specifically showed that 
Unfortunately, most gradual systems with run-time semantics
(as opposed to type erasure as in TypeScript) typically
imposes significant run-time performance overheads to provide these
semantics.  This has lead to a significant body of research,
on one hand developing techniques to optimise gradual typing 
\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
and on the other, developing techniques to evaluate the
resulting performance  \cite{Takikawa2016,Greenman2019jfp}.

This paper builds upon our recent work on optimising transient
type checks \cite{Roberts2017,roberts-and-co-ecoop-2019} by
investigating how benchmark results can be repurposed to identify which
particular transient type checks are resistant to optimisation.

Specifically, the contributions of this work are:

\begin{itemize}
  \item an approach to identifying gradual type annotations
        that cause significant performance effects
  % SM: after discussion with Isaac, this was done
  %     by some Reticulated Python paper before
  % and Isaac's adaptation is restricted to using rounding rules
  %    (>= 0.5 is round up) that give useful results (at least 1 type check)
  %    for selecting the number of type checks
  % \item adapting the Takikawa protocol for languages and implementations
  %       where performance effects are dominated by type annotations instead
  %       of transitions between typed and untyped modules
  \item an observation that the overhead of Moth's transient type checking on small benchmarks (with 10--250 type annotations) is usually caused by only one or two type annotations
\end{itemize}

@@

To identify type annotations causing anomalies,
% To investigate Moth's support for transient type checks further, 
we adapted the Takikawa protocol
and took inspiration from the Takikawa-Vitousek variant.
For each benchmark, we generated 100 partially typed versions,
or fewer if the benchmark has less than 11 types.
We did an even split so that for each $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations. We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to randomly choose the type annotations each configuration contained, and we ensured that no duplicate configurations were generated. In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark (or 97 in the case of our Storage benchmark, since it only has 10 type annotations).

@@

% In our previous work , we evaluated the performance of transient type checks using 21 benchmarks. We started with the 14 benchmarks of the : Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks. We added another \ugh{nine}\sm{14 plus 9 isn't 21} benchmarks taken from the gradual-typing literature.\sm{it's also unclear whether this is about previous work or what we do now.}
% Our goal was to complement the \AWFY with additional ones that are used for similar experiments and can be ported to Grace.
% We surveyed a number of papers
% \citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
% and selected additional benchmarks that have been used by multiple papers.
% Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
% or were available in different versions.
% While not always behaviourally equivalent,
% we chose the \AWFY versions since we already used them to
% establish a performance baseline.
% The selected benchmarks as well as the papers in which they were used are shown in
% Table~\ref{tab:gradual-benchmarks}.
% \sm{this paragraph needs to be cleaned up.
% We need to say what we use now. Clearly, and not intermingled with previous work.
% And then, afterwards we can say whether this is the same we did before. or how it differs.
% }
%
% \begin{table}[htb]
%   \caption{Benchmarks selected from literature.}
%   \label{tab:gradual-benchmarks}
%   \begin{center}
%     \begin{tabular}{l l r}
%       Fannkuch & \cite{Vitousek2017,Greenman2018} \\
%       Float & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
%       Go & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
%       NBody & \cite{Kuhlenschmidt:2018:preprint,Vitousek2017,Greenman2018} & used \cite{Marr2016} \\
%       Queens & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} & used \cite{Marr2016} \\
%       PyStone & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
%       Sieve & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} & used \cite{Marr2016} \\
%       Snake & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} \\
%       SpectralNorm & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
%     \end{tabular}
%   \end{center}
% \end{table}

Each benchmark has a complete set of type annotations,
\ie, all possible type annotations are present,
which is verified in the parser. \isaac{This is misleading, in a few cases, some of the type annotations are in fact "Unknown", which is exactly what my "removing" of type annotations places}

@@

We used the same machine, version of Moth and Graal as previously, specifically all experiments were executed on a machine running Ubuntu Linux 16.04.4, with Kernel 3.13. The machine has two Intel Xeon E5-2620 v3 2.40GHz, with 6 cores each, for a total of 24 hyperthreads.
We also used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43. Benchmarks were executed one by one to avoid interference between them. The analysis of the results and plots where generated using Python 3.7.3 and PGFPLOTS 1.16. To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, are available online.\footnote{\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}

@@

Of particular note is that some of the graphs (Permute, Storage, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly-horizontal lines. Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away. List appears to have three performance modes: the graph consists of three mostly flat lines, at about 1 times slower, 1.5 times slower, and 1.8 times slower than untyped code.

@@

The column graphs, right of the corresponding scatter plot, shows the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code). The y-axes for the column graphs are the same as the associated scatter graphs. However, though the right y-axes for the column graphs appear to be identical to the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.

@@

For each benchmark we highlighted one or two type annotations that appear to show a pattern for the typing lattice performance scatter plots, or in the case of SpectralNorm and Json had higher than usual columns. The identified types are those represented by the red and blue columns. The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation present, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither. Though we exhaustively inspected such colour coded scatter plots for all 1,775 type annotations across all 21 benchmarks, the only patterns we noticed are those shown in \ref{f:pattern}.

As can be seen, most of the time the patterns we found in the typing lattice graphs correspond to outliers in the single type annotation column graphs. For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs. For Permute there is actually a significant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red. DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect does not appear to be cumulative, however: the purple dots are about the same height as the red and blue ones, so \emph{either} annotation is sufficient for the full benefit.

List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present. Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotation is present, but not the blue one. We had previously identified this benchmark and the aforementioned type annotations \cite{roberts-and-co-ecoop-2019} as being the worst case for Moth.

@@

In particular, observe that the grey dots for Snake, Towers, DeltaBlue, Go, and List are mostly flat horizontal lines, this indicates that by simply deleting the red and blue type annotations, the performance impact of transient type checking becomes negligible. However for CD and Richards, the transient type checking overhead of the grey dots appears linear, albeit still less than with the red and blue type annotations present. Thus we can observe that usually, only a couple of type annotations are responsible for the overhead caused by Moth's transient typed checking, whereas the rest are ``free''.

@@

In this paper we have investigated how benchmarks can be
repurposed to determine precisely which transient type checks are
resistant to the optimisations provided by a just-in-time compiling
virtual machine.  We observed that many of our benchmark results
conducted under the Takikawa protocol 
seemed to have a characteristic bimodal performance profile: some
of the benchmark configurations ran significantly slower than
the remaining configurations. We also observed trimodal profiles, as well as performance increases when type annotations where added.

By inspecting graphs of the performance of where only one type annotation is present, we can easily identify type annotations that likely have a significant effect on performance. By then inspecting the typing lattice graphs colour coded based on whether such type annotation was present or not, we were often able to notice patterns.
In particular, these patterns suggested the just one or two type annotations are likely responsible for the bimodal performance and most of the overhead caused by Moth's transient type checking.
Though every type annotation that appeared to have a significant effect across the typing lattice also showed a significant performance effect when no other typing annotations where present, the converse did not hold.

This is preliminary work, in particular we have not yet identified exactly why these type annotations appear to have an effect on performance.
There are also a number of threats to validity. 
Regarding construct validity, 
our underlying
implementation may contain undetected bugs that affect the semantics
or performance of the gradual typing checks. Regarding internal
validity,
our benchmarking harness runs on the same implementation
and therefore is subject to the same issues.
Regarding external validity, 
%\RR{The next two sentences seem to be in the wrong place (they state an advantage rather than a threat); maybe move them into the previous paragraph?}
%\kjx{I'm happy with them here about external validity}
Moth is built
on the Truffle and Graal toolchain, so we expect
to resemble other Graal
VMs doing similar AST-based optimizations of transient type checks.
Because we rely on common techniques, 
we expect our results to be transferable to other JIT implementations as well.

@@

>>>>

\setcopyright{acmlicensed}
\acmPrice{15.00}
\acmDOI{10.1145/3358504.3361232}
\acmYear{2019}
\copyrightyear{2019}
\acmISBN{978-1-4503-6987-9/19/10}
\acmConference[VMIL '19]{Proceedings of the 11th ACM SIGPLAN International Workshop on Virtual Machines and Intermediate Languages}{October 22, 2019}{Athens, Greece}
\acmBooktitle{Proceedings of the 11th ACM SIGPLAN International Workshop on Virtual Machines and Intermediate Languages (VMIL '19), October 22, 2019, Athens, Greece}

@@

\section{Introduction}
Gradual typing aims to add static type annotations to dynamic languages, increasing their safety while maintaining flexibility \citep{GiladPluggable2004,Siek2006,XXXSiek2015}, and/or, permitting dynamic type annotations within static languages, increasing flexibility whilst maintaining some safety \citep{AbadiTOPLAS1991}.

There is a vast spectrum of different approaches to gradual typing \cite{kafka18,bensurvey18icfp}. Here we measure the performance of ``transient'' or ``type-tag'' checks (as in Reticulated Python), which offer first-order semantics: they check that an object's type constructor or names of supported methods match any static types that the object flows through, but not the return/argument types of such methods \cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.

Unfortunately most gradual systems with run-time semantics (as opposed to type erasure as in TypeScript) typically impose significant run-time performance overheads. This has lead to a significant body of research to develop techniques to optimise gradual typing \citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}.

This has also lead to a technique to evaluate the performance of gradual typing, such as the so called so-called ``Takikawa'' protocol \cite{Takikawa2016,Greenman2019jfp,vitousek-transient-arXive-2019}. The Takikawa protocol was created to measure the runtime cost of gradual typing by testing various configurations of typed/untyped code. This approach was designed to investigate the the performance effect of the amount of typed vs untyped code, in particular it shows that how simply adding more type annotations does not always produce a uniform effect on performance. Here we adapt their approach in order to identify individual type annotations that may be responsible for performance effects.

This paper builds upon our recent work on optimising transient type checks \cite{Roberts2017,roberts-and-co-ecoop-2019} to make the following contributions:
\begin{itemize}
  \item an approach to identifying individual gradual type annotations
        that cause significant performance effects
  % SM: after discussion with Isaac, this was done
  %     by some Reticulated Python paper before
  % and Isaac's adaptation is restricted to using rounding rules
  %    (>= 0.5 is round up) that give useful results (at least 1 type check)
  %    for selecting the number of type checks
  % \item adapting the Takikawa protocol for languages and implementations
  %       where performance effects are dominated by type annotations instead
  %       of transitions between typed and untyped modules
  \item an observation that the overhead of Moth's transient type checking on small benchmarks (with 10--250 type annotations) is usually caused by only one or two type annotations
\end{itemize}

@@

Unlike such prior work, we wish to identify which type annotations cause anomalies,
% To investigate Moth's support for transient type checks further, 
thus we adapted the Takikawa protocol
and took inspiration from the Takikawa-Vitousek variant.
For each benchmark, we generated 100 partially typed versions,
or fewer if the benchmark has less than 11 types.
We did an even split so that for each $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations. We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to randomly choose the type annotations each configuration contained, and we ensured that no duplicate configurations were generated. In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark (or 97 in the case of our Storage benchmark, since it only has 10 type annotations).

@@

% In our previous work , we evaluated the performance of transient type checks using 21 benchmarks. We started with %the 14 benchmarks of the : Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks. We added another \ugh{nine}\sm{14 plus 9 isn't 21} benchmarks taken from the gradual-typing literature.\sm{it's also unclear whether this is about previous work or what we do now.}
% Our goal was to complement the \AWFY with additional ones that are used for similar experiments and can be ported to Grace.
% We surveyed a number of papers
% \citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
% and selected additional benchmarks that have been used by multiple papers.
% Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
% or were available in different versions.
% While not always behaviourally equivalent,
% we chose the \AWFY versions since we already used them to
% establish a performance baseline.
% The selected benchmarks as well as the papers in which they were used are shown in
% Table~\ref{tab:gradual-benchmarks}.
% \sm{this paragraph needs to be cleaned up.
% We need to say what we use now. Clearly, and not intermingled with previous work.
% And then, afterwards we can say whether this is the same we did before. or how it differs.
% }
%
% \begin{table}[htb]
%   \caption{Benchmarks selected from literature.}
%   \label{tab:gradual-benchmarks}
%   \begin{center}
%     \begin{tabular}{l l r}
%       Fannkuch & \cite{Vitousek2017,Greenman2018} \\
%       Float & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
%       Go & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
%       NBody & \cite{Kuhlenschmidt:2018:preprint,Vitousek2017,Greenman2018} & used \cite{Marr2016} \\
%       Queens & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} & used \cite{Marr2016} \\
%       PyStone & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
%       Sieve & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} & used \cite{Marr2016} \\
%       Snake & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} \\
%       SpectralNorm & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
%     \end{tabular}
%   \end{center}
% \end{table}

Each benchmark has an almost complete set of type annotations,
\ie, static type annotations are present at almost all possible points.
%ISAAC: I changed this as it was misleading, in a few cases, some of the type annotations are in fact "Unknown", which is exactly what my "removing" of type annotations places

@@

All our experiments used the same machine, Graal, and Moth as previously; the machine has two Intel Xeon E5-2620 v3 2.40GHz, with 6 cores each, for a total of 24 hyperthreads.
The machine was running Ubuntu Linux 16.04.6, with Kernel 4.4, and we used ReBench 1.0 \citep{ReBench:2018} and Java 1.8.0\_191 Graal 0.43. Benchmarks were executed one by one to avoid interference between them. The analysis of the results and plots where generated using Python 3.7.3 and PGFPLOTS 1.16. To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, are available online.\footnote{\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}

In our previous work we also compared the performance of untyped code on Moth against state-of-the-art VMs: Java, Node.js using the V8 JavaScript VM, and the Higgs JavaScript VM. Java was the fastest of these, and on average V8 was about 1.8x slower than Java, Moth was 2.3x slower, and Higgs was 10.4x slower.
Hence we continue to argue that ``Moth is a suitable platform to assess the impact of our approach to gradual type checking, because its performance is close enough to state-of-the-art VMs, and run-time overhead is not hidden by slow baseline performance.''

@@

Of particular note is that some of the graphs (Permute, Storage, Towers, and Richards) show bimodal performance profiles, that is, two separate roughly-horizontal lines. Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away. List shows three performance modes: the graph consists of three mostly flat lines, at about 1 times slower, 1.5 times slower, and 1.8 times slower than untyped code.

@@

The column graphs, right of the corresponding scatter plot, shows the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code). The y-axes for the column graphs are the same as the associated scatter graphs.

@@

For each benchmark we highlighted one or two type annotations that seem to show a pattern for the typing lattice performance scatter plots, or in the case of SpectralNorm and Json had higher than usual columns. The identified types are those represented by the red and blue columns. The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation present, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither. Though we exhaustively inspected such colour coded scatter plots for all 1,775 type annotations across all 21 benchmarks, the only patterns we noticed are those shown in \ref{f:pattern}.

As can be seen, most of the time the patterns we found in the typing lattice graphs correspond to outliers in the single type annotation column graphs. For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears responsible for the upper/slowest half of the typing lattice graphs. For Permute there is actually a significant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red. DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect is not cumulative, however: the purple dots are about the same height as the red and blue ones, so \emph{either} annotation is sufficient for the full benefit.

List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present. Richards is particularly odd as it shows a performance decrease that occurs when the red type annotation is present, but not the blue one. We had previously identified this benchmark and the aforementioned type annotations \cite{roberts-and-co-ecoop-2019} as being the worst case for Moth.

@@

In particular, observe that the grey dots for Snake, Towers, DeltaBlue, Go, and List are mostly flat horizontal lines, this indicates that by simply deleting the red and blue type annotations, the performance impact of transient type checking becomes negligible. However for CD and Richards, the transient type checking overhead of the grey dots is roughly linear, albeit still less than with the red and blue type annotations present. Thus we can observe that usually, only a couple of type annotations are responsible for the overhead caused by Moth's transient typed checking, whereas the rest are ``free''.

@@

In this paper we have investigated how benchmarks can be
repurposed to determine precisely which transient type checks are
likely resistant to the optimisations provided by a just-in-time compiling
virtual machine.
However, detailed analysis will be needed in order to identify exactly what causes such performance effects, such as that we previously did for the List benchmark \cite{roberts-and-co-ecoop-2019}.
We observed that many of our benchmark results
conducted under the Takikawa protocol 
showed a characteristic bimodal performance profile: some
of the benchmark configurations ran significantly slower than
the remaining configurations. We also observed trimodal profiles, as well as performance increases when type annotations where added.

By inspecting graphs of the performance of where only one type annotation is present, we can easily identify type annotations that likely have a significant effect on performance. By then inspecting the typing lattice graphs colour coded based on whether such type annotation was present or not, we were often able to notice patterns.
In particular, these patterns suggested the just one or two type annotations are likely responsible for the bimodal performance and most of the overhead caused by Moth's transient type checking.
Though every type annotation that showed a significant effect across the typing lattice also showed a significant performance effect when no other typing annotations where present, the converse did not hold.

This is preliminary work, in particular we have not yet identified exactly why these type annotations show an effect on performance.
There are also a number of threats to validity. 
Regarding construct validity, 
our underlying
implementation may contain undetected bugs that affect the semantics
or performance of the gradual typing checks. Regarding internal
validity,
our benchmarking harness runs on the same implementation
and therefore is subject to the same issues.
Regarding external validity, 
%\RR{The next two sentences seem to be in the wrong place (they state an advantage rather than a threat); maybe move them into the previous paragraph?}
%\kjx{I'm happy with them here about external validity}
Moth is built
on the Truffle and Graal toolchain, so we expect
to resemble other Graal
VMs doing similar AST-based optimizations of transient type checks.
Because we rely on common techniques, 
we expect our results to be transferable to other JIT implementations as well.

@@

\begin{acks}
The authors would like to thank the anonymous reviewers for their
valuable comments and helpful suggestions. This work is supported
in part by the \grantsponsor{Marsden Fund}{Royal Society of New Zealand (Te Ap\={a}rangi) Marsden Fund (Te P\={u}tea Rangahau a Marsden)}{https://royalsociety.org.nz/what-we-do/funds-and-opportunities/marsden/} under grant \grantnum{Marsden Fund}{VUW1815}.
\end{acks}
